---
title: "Random Num Generation HW6"
author: "Yichu Li.  "
date: "10/16/2018"
output:
  pdf_document: default
  html_document: default
papersize: letter
fontsize: 11pt
abstract: |
  In the project, we want to create a process that simulates observations from a complex density function, and practice rejection sampling method. Also, apply the method to Beta Mixture Model. 
---



# 5.2.1 Rejection sampling
##  Show that g is a mixture of Gamma distributions.
 First calculate the value of integral. The integral is in the form of a mixture of Gamma distributions with $\alpha = \theta, \theta+1/2$ and $\beta = 1, 1$ correspondingly. According to the property of the Gamma distribution, 
\begin{align}
      &\int_0^{\infty}(2x^{\theta - 1} + x^{\theta-1/2}) e^{- x} dx
  = 2 \int_0^{\infty}x^{\theta - 1} e^{- x} dx + \int_0^{\infty}x^{\theta-1/2} e^{-x} dx\\ \\
  = 2 &\Gamma(\theta) + \Gamma(\theta + 1/2)
 \end{align}
According to the property of the Gamma distribution,
$$\int_0^\infty x^{\alpha - 1} e^{-\beta / \alpha} dx 
= \beta^\alpha \Gamma(\alpha)$$
The value of the constant $C$ is
$$C = \frac{1}{2\Gamma(\theta) + \Gamma(\theta + 1/2)}$$
Since $g(x)$ is a probability density on $(0, \infty)$
\begin{align}
g(x) & = \frac{1}{2\Gamma(\theta) + \Gamma(\theta + 1/2)}
(2x^{\theta - 1} + x^{\theta-1/2}) e^{- x}\\ \\
g(x) & = \frac{1}{2\Gamma(\theta) + \Gamma(\theta + 1/2)}
2x^{\theta - 1} e^{- x} + x^{\theta-1/2} e^{- x}\\ \\
g(x) & = \frac {2 \Gamma(\theta)}{2\Gamma(\theta) + \Gamma(\theta + 1/2)} \frac{2x^{\theta - 1} e^{- x}}{\Gamma(\theta)}
        +\frac {\Gamma(\theta + 1/2)}
        {2\Gamma(\theta) + \Gamma(\theta + 1/2)}
        \frac{x^{\theta-1/2} e^{- x}}{\Gamma(\theta + 1/2)}\\ \\
 \end{align}



## Design a procedure (pseudo-code) to sample from g(x)
To sample from $g(x)$, we first generate random numbers $U$ for weights from the standard uniform distribution $U(0, 1)$ for $10,000$ times as required. Then, compare the values with $w_1$. If $U < w_1$, return $X \sim Gamma(\theta, 1)$; otherwise, return $X \sim Gamma(\theta + 1/2, 1)$.\
We chose $\theta = 2$ 
```{r}
N <- 10000                
U <- runif(N)
rand.samples <- rep(NA,N)

theta <- 2
w1 <- 2*gamma(theta) / (2*gamma(theta) + gamma(theta + 0.5))
C <- 1/(2*gamma(theta) + gamma(theta + 0.5))

for(i in 1:N){
  if(U[i] < w1){
    rand.samples[i] <- rgamma(1,theta,1)
  }
  else{
    rand.samples[i] <- rgamma(1,theta + 0.5,1)
  }
}
plot(density(rand.samples), main = "Density Estimate of the MG distribution")

x = seq(0, 40, .1)
truth = w1*dgamma(x, theta, 1) + (1-w1) * dgamma(x,theta + 0.5, 1)
plot(density(rand.samples), 
     main = "Density Estimate of the Mixture Model",ylim = c(0,.4),lwd = 2)
lines(x, truth, col = "red",lwd = 2)

legend("topleft", c("Density","Estimated Density"), 
       col = c("red","black"),cex = 1,lwd = 1)
```

##  Design a procedure (pseudo-code) to use rejection sampling to sample from f using  g as the instrumental distribution.
We let $f(x) = \frac{q(x)}{C_1}$, 
and $q(x) = \sqrt{x + 4} x^{\theta-1} e^{-x}$\
First, find the value of $\alpha$
$$\alpha = sup \frac {q(x)} {g(x)}
         = sup \frac {\sqrt{x + 4} x^{\theta-1} e^{-x}} 
         {C(2x^{\theta - 1} + x^{\theta-1/2}) e^{- x}}
         = sup \frac {\sqrt{x + 4}} {C (2 + x^{1/2})}$$
Let $y = \frac {\sqrt{x + 4}} {C (2 + x^{1/2})}$ and $y' = 0$.
Solve it and we get $x = 4$ 
Thus, 
$$\alpha = \frac {\sqrt2} {2C}$$
First sample $X \sim g(x)$ and $U \sim U(0, 1)$. Then compare the values of $U$ and $\frac{q(x)}{\alpha g(x)}$
```{r}

N <- 1000
samplef <- rep(NA,N)
theta <- 2
w1 <- 2 * gamma(theta) / (2 * gamma(theta) + gamma(theta + 0.5))
C <- 1 / (2 * gamma(theta) + gamma(theta + 0.5))
alpha <- sqrt(2) / (2 * C)

for(k in 1:N){
  V <- runif(1)
  if(V < w1){
    X <- rgamma(1,theta,1)
  }
  else{
    X <- rgamma(1,theta+0.5,1)
  }
  U <- runif(1)
  q_x <- sqrt(4 + X) * X^(theta - 1) * exp(-1 * X)
  g_x <- C * (2 * X ^ (theta - 1) + X ^ (theta - 0.5)) * exp(-1 * X)
  m <- q_x/(alpha * g_x)
  if (U > m){cat("one more iteration"); k <- k-1; next}
  else samplef[k] <- X
  }

plot(density(samplef),main = "Density Estimate of f")
```

#  5.2.2 Mixture Proposal
##  Design a procedure (pseudo-code) to sample from f using a mixture of Beta distributions
Define $g(x)$ as a mixture of 2 Beta distribution $beta(\theta, 1)$ and $beta(1, \beta)$. Assume they have the weights of $p_1$ and $p_2$, $p_1 + p_2 = 1$. Thus,
\begin{align}
g(x)  &= p_1 \frac {x^{\theta - 1}} {beta(\theta, 1)}
        +p_2 \frac {(1 - x)^{\beta - 1}} {beta(1, \beta)}\\ \\
\alpha &= sup \frac {q(x)} {g(x)}
 \end{align}

The coefficient of $x^{2\theta - 3}$ gives the equation  
\begin{align}
\frac{(\theta - 1)(1 + x^2) - 2 x^2}{1 + x^2}
                    \frac{p_1}{beta(\theta, 1)}
&= \frac{1}{1 + x^2} (\theta - 1) \frac{p_1}{beta(\theta, 1)}\\ \\
(\theta - 1) - \frac{2 x^2}{1 + x^2}
&= \theta-1
\end{align}
which has $x = 0$ as the only solution.\

The coefficient of $x^{2\beta - 3}$ gives the equation  
\begin{align}
[\frac{x (1 - x)}{\sqrt{2 + x^2}} - (\beta - 1) \sqrt{2 + x^2}]
\frac{p_2}{beta(1, \beta)}   
&= -\sqrt{2 + x^2} (\beta - 1) \frac{p_2}{beta(1, \beta)}\\ \\
x (1 - x)
&= 0
\end{align}
which has $x = 0, 1$ as the solutions.\
Now take $x = 0$ back into the original numerator, it becomes
$$
[- (\beta - 1) \sqrt{2}] \frac{p_2}{beta(1, \beta)} 
- \sqrt{2} [-(\beta -1) \frac{p_2}{beta(1, \beta)}] = 0
$$
Therefore, $x = 0$ is the point when the supremum is reached.\
In our model, we let $(\theta, \beta) = (3, 4)$, which gives $p_1 = p_2 = 0.5$. In this way,
$$\alpha = \frac{q(0)}{g(0)} 
         = \frac{\sqrt{2} beta(1, \beta)}{p_2}= 0.7071$$
Compare the values of $U$ and $\frac{q(x)}{\alpha g(x)}$: If U is bigger, return to the first step; return X otherwise.
The returned value is a random sample from the density function $f(x)$.
```{r}
#Plotting 
x <- seq(0,1,.001)
p1 <- 0.5
theta <- 3
beta <- 4
denomina <- p1*(1 / beta(theta,1)) * x ^ (theta-1) +
  (1-p1) * (1/beta(1,beta)) * ((1-x)^(beta-1))

nomina <- x^(theta - 1) / (1 + x^2) + sqrt(2 + x^2)*(1 - x)^(beta-1)
test <- nomina/denomina

plot(x, test, main = "upper limit", lwd = 2)
X <- 0
denomina <- p1 * (1 / beta(theta,1)) * X^(theta-1) + 
  (1-p1) * (1/beta(1, beta)) * ((1-X)^(beta-1))
nomina <- X^(theta-1) / (1 + X^2) + sqrt(2 + X^2) * (1 - X)^(beta - 1)
alpha <- nomina/denomina
N <- 1000
samplef <- rep(NA,N)
k <- 1
while(k <= N){
  V <- runif(1)
  if(V < p1){
    X <- rbeta(1, theta, 1)
  }
  else{
    X <- rbeta(1, 1, beta)
  }
  U <- runif(1)
  q_x <- X^(theta - 1) / (1 + X^2) + sqrt(2 + X^2) * (1-X)^(beta-1)
  g_x <- p1 * (1/beta(theta,1)) * X^(theta-1) 
         + (1-p1) * (1/beta(1,beta)) * ((1-X)^(beta-1))
  m <- q_x / (alpha * g_x)
  if ( U > m ) {next}
  samplef[k] <- X
  k <- k + 1
 
}

plot(density(samplef), main = "Density Estimate of f")
```


##  Design a procedure dealing with the two components separately using individual Beta distributions.
We let $q_1(x) = \frac{x^{\theta - 1}}{1 + x^2}$, $g_1(x) \sim beta(\theta, 1)$, then we have
$$\alpha_1 = sup \frac{q_1(x)}{g_1(x)} = \frac{beta(\theta, 1)}{1 + x^2}$$
at $x = 0$, $\alpha_1 = beta(\theta, 1)$.\
Similarly for $q_2(x) =\sqrt{2 + x^2} (1-x)^{\beta - 1}$, $g_2(x) \sim beta(1, \beta)$
$$\alpha_2 = sup \frac{q_2(x)}{g_2(x)} = beta(1, \beta) \sqrt{2+x^2}$$
at $x = 1$, $\alpha_2 = \sqrt{3} beta(1, \beta)$\
For the pseudo-code, we first sample $k$ from $\{1, 2\}$ with probabilities $\{\frac{\alpha_1}{\alpha_1+\alpha_2}, \frac{\alpha_2}{\alpha_1+\alpha_2}\}$. 
Then, we generate $V$ from $U (0, 1)$. 
Third, compare the values of $V$ and$\frac{q_k(x)}{\alpha g_k(x)}$.
If V is bigger, reject.
```{r}
N <- 1000
samplef <- rep(NA,N)
theta <- 3
beta <- 4
alpha_1 <- beta(theta,1)
alpha_2 <- sqrt(3)*beta(1,beta)
w1 <- alpha_1 / (alpha_1 + alpha_2)
k <- 1
while(k <= N){
  #sampling from g
  V <- runif(1)
  if(V < w1){
    #k=1
    X <- rbeta(1,theta,1)
    U <- runif(1)
    q_x <- X^(theta-1) / (1 + X^2)
    g_x <- X^(theta-1) / beta(theta,1)
    m <- q_x / (alpha_1 * g_x)
  }
  else{
    #k=2
    X <- rbeta(1,1,beta)
    U <- runif(1)
    q_x <- sqrt(2 + X^2)*(1 - X)^(beta - 1)
    g_x <- (1 - X)^(beta - 1) / beta(1, beta)
    m <- q_x / (alpha_2 * g_x)
  }
  if ( U > m ) { next }
  samplef[k] <- X
  k <- k+1 
  }

plot(density(samplef), main = "Density Estimate of f generated from rejection sampling")
```

